### 2013-03-25 doc/dissertation/document/chapters/mc_background/mc_background.tex
some cleanup in method comparisions in mc background

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Method Selection}
The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
to estimate the error in residual Monte Carlo solve instead of the
direct method outlined in \S~\ref{sec:direct_mc}. To demonstrate
the effectiveness of the adjoint method over the direct method within
the context of MCSA, we choose the 2D time-dependent Poisson equation
as a simple model problem:
\begin{equation}
  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of grid size, fixing the spectral radius of the
system at a constant value for each variation. Both the five-point and
nine-point stencils were used with both the direct and adjoint
solvers. For each case, $N \times N$ total random walk permutations
were computed per MCSA iteration where $N \times N$ is the number of
discrete grid points in the system. Solver parameters were set to a
weight cutoff of \sn{1}{-4} for the stochastic linear solver and a
convergence tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance as a function of the problem size. All
computations presented in this section and
\S~\ref{subsec:sequential_comparison} were completed on a 3.0 GHz
Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we
compare its performance to Halton's Sequential Monte Carlo method on
which our previous work in this area was based. For this comparison,
we use the same transient Poisson problem as described in the previous
section and choose only the 5-point stencil to discretize the
Laplacian operator as the previous results yielded little qualitative
difference between the discretizations. Both MCSA and Halton's method
are used with the adjoint Monte Carlo solver. In order to complete the
same study as in the previous section, the number of histories
computed by the Monte Carlo solver at each iteration had to be doubled
to $2 \times N \times N$ in order to ensure convergence in Sequential
Monte Carlo Method. Figure~\ref{fig:seq_cpu_time} gives the CPU time
results for this comparison as a function of problem size while
Figure~\ref{fig:seq_iterations} gives the number of iterations to
converge as a function of problem size. In both cases, using the Monte
Carlo solver as a synthetic acceleration rather than in a pure
residual Monte Carlo scheme resulted in a reduction in both CPU time
and iterations required to converge. The additional Richardson
extrapolation between each Monte Carlo solve in the MCSA method gives
a better converged residual source to use with the Monte Carlo
calculation while the Sequential method requires more iterations to
achieve the same level of convergence in the residual.

>>>>

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Method Selection}
The MCSA method defined in Eq.~(\ref{eq:mcsa}) uses the adjoint method
to estimate the error in a residual Monte Carlo solve instead of the
direct method outlined in \S\ref{sec:direct_mc}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we choose the two-dimensional time-dependent Poisson
equation as a simple model problem:
\begin{equation}
  \frac{\partial \ve{u}}{\partial t} = \nabla^2 \ve{u}\:.
  \label{eq:poisson_equation}
\end{equation}
For all comparisons, a single time step is computed with backwards Euler time
integration. The Laplacian is differenced on a square Cartesian grid with a
second-order five-point stencil,
\begin{equation}
  \nabla^2_5 = \frac{1}{\Delta^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]\:,
  \label{eq:five_point_stencil}
\end{equation}
and a fourth-order nine-point stencil,
\begin{multline}
  \nabla^2_9 = \frac{1}{6\Delta^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4
    u_{i,j-1} + 4 u_{i,j+1} + u_{i-1,j-1}\\ + u_{i-1,j+1} +
    u_{i+1,j-1} + u_{i+1,j+1} - 20 u_{i,j}]\:,
  \label{eq:nine_point_stencil}
\end{multline}
both assuming a grid size of $\Delta$ in both the $i$ and $j$ directions. For
a single time step solution, we then have the following sparse linear system
to be solved with the MCSA method:
\begin{equation}
  \ve{A} \ve{u}^{n+1} = \ve{u}^n\:.
  \label{eq:poisson_eq_lin_sys}
\end{equation}
Both the stencils will be used to vary the size and density of the sparse
linear system in Eq.~(\ref{eq:poisson_eq_lin_sys}).

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant $\Delta$ was used
with varying values of the number of mesh elements, fixing the
spectral radius of the system at a constant value for each
variation. Both the five-point and nine-point stencils were used with
both the direct and adjoint solvers. For each case, $N \times N$ total
random walk permutations were computed per MCSA iteration where $N
\times N$ is the number of discrete grid points in the system. Solver
parameters were set to a weight cutoff of \sn{1}{-4} for the
stochastic linear solver and a convergence tolerance of \sn{1}{-8} for
the MCSA iterative solver.  Figure~\ref{fig:poisson_cpu_time} gives
the CPU time needed for each case to converge in seconds and
Figure~\ref{fig:poisson_iterations} gives the number of iterations
needed for each case to converge to the specified tolerance as a
function of the problem size. All computations presented in this
section and \S\ref{subsec:sequential_comparison} were completed on a
3.0 GHz Intel Core 2 Quad Q9650 CPU machine with 16 GB 1067 MHz DDR3
memory.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Sequential Monte Carlo}
\label{subsec:sequential_comparison}
To further motivate using Monte Carlo Synthetic Acceleration, we
compare its performance to Halton's Sequential Monte Carlo method on
which previous work in this area was based. For this comparison, we
use the same transient Poisson problem as described in the previous
section and choose only the 5-point stencil to discretize the
Laplacian operator as the previous results yielded little qualitative
difference between the discretizations. Both MCSA and Halton's method
are used with the adjoint Monte Carlo solver. In order to complete the
same study as in the previous section, the number of histories
computed by the Monte Carlo solver at each iteration had to be doubled
to $2 \times N \times N$ in order to ensure convergence in Sequential
Monte Carlo Method. For the majority of the problems in the previous
section, the Sequential method used with $N \times N$ histories would
not converge. Figure~\ref{fig:seq_cpu_time} gives the CPU time results
for this comparison as a function of problem size while
Figure~\ref{fig:seq_iterations} gives the number of iterations to
converge as a function of problem size. In both cases, using the Monte
Carlo solver as a synthetic acceleration rather than in a pure
residual Monte Carlo scheme resulted in a reduction in both CPU time
and iterations required to converge. The additional Richardson
extrapolation between each Monte Carlo solve in the MCSA method gives
a better converged residual source to use with the Monte Carlo
calculation while the Sequential method requires more iterations to
achieve the same level of convergence in the residual.
